---
title: "Climate Similarity using Koppen-Geiger Classification"
author: "Ana"
date: "2026-03-25"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
---

## Setup

Install any missing packages first (only need to run once):

```{r install-packages, eval=FALSE}
install.packages(c("terra", "sf", "rnaturalearth", "rnaturalearthdata",
                    "exactextractr", "dplyr", "tidyr", "data.table"))
```

Load libraries:

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())

library(terra)
library(sf)
library(rnaturalearth)
library(exactextractr)
library(dplyr)
library(tidyr)
library(data.table)
```

Set working directory to the project folder:

```{r set-wd}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Step 1: Load Raster and World Map

We use the Koppen-Geiger climate classification raster from Beck et al. (2023), which classifies every pixel on Earth into one of 30 climate zones. We use the 1991--2020 period at 0.1 degree resolution (~11km per pixel).

The world country polygons come from `ne_countries` in the `rnaturalearth` package (same shapefile used by the rest of the team for consistency).

**Data source:** Beck, H. E. et al. (2023). High-resolution (1 km) Koppen-Geiger maps for 1901--2099 based on constrained CMIP6 projections. *Scientific Data*. Downloaded from https://www.gloh2o.org/koppen/

```{r load-data}
koppen_raster <- rast("../Data/Raw/1991_2020/koppen_geiger_0p1.tif")
koppen_raster

world <- ne_countries(scale = "medium", returnclass = "sf")
```

The raster values are integers 1--30, each corresponding to a Koppen-Geiger climate class:

```{r legend}
koppen_legend <- data.frame(
  value = 1:30,
  code = c("Af","Am","Aw","BWh","BWk","BSh","BSk","Csa","Csb","Csc",
           "Cwa","Cwb","Cwc","Cfa","Cfb","Cfc","Dsa","Dsb","Dsc","Dsd",
           "Dwa","Dwb","Dwc","Dwd","Dfa","Dfb","Dfc","Dfd","ET","EF"),
  description = c("Tropical rainforest","Tropical monsoon","Tropical savannah",
                   "Arid desert hot","Arid desert cold","Arid steppe hot","Arid steppe cold",
                   "Temperate dry summer hot","Temperate dry summer warm","Temperate dry summer cold",
                   "Temperate dry winter hot","Temperate dry winter warm","Temperate dry winter cold",
                   "Temperate no dry hot","Temperate no dry warm","Temperate no dry cold",
                   "Cold dry summer hot","Cold dry summer warm","Cold dry summer cold","Cold dry summer very cold winter",
                   "Cold dry winter hot","Cold dry winter warm","Cold dry winter cold","Cold dry winter very cold winter",
                   "Cold no dry hot","Cold no dry warm","Cold no dry cold","Cold no dry very cold winter",
                   "Polar tundra","Polar frost")
)
knitr::kable(koppen_legend, caption = "Koppen-Geiger Climate Classification Legend")
```

Quick visual check of the raster:

```{r plot-raster, fig.width=10, fig.height=5}
plot(koppen_raster, main = "Koppen-Geiger Climate Zones (1991-2020)")
```

## Step 2: Extract Climate Zones Per Country

We overlay each country polygon on the raster using `exact_extract()`. This returns every raster pixel that falls inside each country, along with a `coverage_fraction` value that accounts for partial pixels at borders.

```{r extract}
climate_extract <- exact_extract(
  koppen_raster,
  world,
  include_cols = "iso_a3"
)

climate_df <- bind_rows(climate_extract)
names(climate_df)[names(climate_df) == "value"] <- "climate_zone"

cat("Total pixels extracted:", nrow(climate_df), "\n")
head(climate_df)
```

## Step 3: Calculate Share of Each Climate Zone Per Country

For each country, we compute the fraction of its total land area that falls in each Koppen zone. We use `coverage_fraction` as the weight so that partial border pixels are handled correctly. Ocean pixels (value = 0) and countries with missing ISO codes are dropped.

```{r shares}
climate_shares <- climate_df %>%
  filter(!is.na(climate_zone), climate_zone != 0) %>%
  group_by(iso_a3, climate_zone) %>%
  summarise(area = sum(coverage_fraction), .groups = "drop") %>%
  group_by(iso_a3) %>%
  mutate(share = area / sum(area)) %>%
  ungroup() %>%
  select(iso_a3, climate_zone, share)

climate_shares <- climate_shares %>%
  filter(!is.na(iso_a3), iso_a3 != "-99")

cat("Countries with climate data:", n_distinct(climate_shares$iso_a3), "\n")
```

**Sanity check:** Spain should be mostly arid steppe cold (BSk = 7) and Mediterranean (Csa = 8, Csb = 9):

```{r check-spain}
spain <- climate_shares %>%
  filter(iso_a3 == "ESP") %>%
  arrange(desc(share)) %>%
  left_join(koppen_legend, by = c("climate_zone" = "value"))
knitr::kable(spain, digits = 3, caption = "Climate zone shares for Spain")
```

## Step 4: Compute Bilateral Climate Similarity

We measure climate similarity between each pair of countries using the **same overlap method** as the Finger-Kreinin index used for food similarity:

$$\text{Climate Similarity}_{ij} = \sum_{z=1}^{30} \min(s_{iz}, s_{jz}) \times 100$$

where $s_{iz}$ is country $i$'s share of land area in climate zone $z$. The score ranges from 0 (completely different climates) to 100 (identical climate distributions).

We pivot the shares into a wide matrix for fast vectorized computation.

```{r similarity}
climate_wide <- climate_shares %>%
  pivot_wider(
    id_cols = iso_a3,
    names_from = climate_zone,
    values_from = share,
    values_fill = 0
  )

country_ids <- climate_wide$iso_a3
climate_mat <- as.matrix(climate_wide[, -1])
rownames(climate_mat) <- country_ids

n <- length(country_ids)
pairs <- combn(n, 2)

cat("Calculating similarity for", ncol(pairs), "country pairs...\n")

similarity_vec <- sapply(seq_len(ncol(pairs)), function(k) {
  sum(pmin(climate_mat[pairs[1, k], ], climate_mat[pairs[2, k], ]))
})

climate_similarity <- data.frame(
  iso3_1 = country_ids[pairs[1, ]],
  iso3_2 = country_ids[pairs[2, ]],
  climate_similarity = similarity_vec * 100
)

cat("Done!", nrow(climate_similarity), "pairs computed\n")
```

## Step 5: Sanity Checks

We verify the results against intuitive expectations. We use a helper function to look up pairs regardless of ordering:

```{r sanity-checks}
lookup_pair <- function(df, c1, c2) {
  df %>% filter(
    (iso3_1 == c1 & iso3_2 == c2) | (iso3_1 == c2 & iso3_2 == c1)
  )
}

cat("Spain & Italy (both Mediterranean, expect HIGH):\n")
lookup_pair(climate_similarity, "ESP", "ITA")

cat("\nNorway & Brazil (expect LOW):\n")
lookup_pair(climate_similarity, "NOR", "BRA")

cat("\nGermany & France (both temperate, expect MODERATE-HIGH):\n")
lookup_pair(climate_similarity, "DEU", "FRA")

cat("\nJapan & USA (both span many zones, expect MODERATE):\n")
lookup_pair(climate_similarity, "JPN", "USA")
```

Top 10 most and least similar country pairs:

```{r top-bottom}
cat("Most similar:\n")
climate_similarity %>% arrange(desc(climate_similarity)) %>% head(10)

cat("\nLeast similar:\n")
climate_similarity %>% arrange(climate_similarity) %>% head(10)
```

Distribution of climate similarity scores:

```{r distribution, fig.width=8, fig.height=4}
hist(climate_similarity$climate_similarity,
     breaks = 50,
     main = "Distribution of Bilateral Climate Similarity Scores",
     xlab = "Climate Similarity (0-100)",
     col = "steelblue",
     border = "white")
abline(v = mean(climate_similarity$climate_similarity), col = "red", lwd = 2, lty = 2)
legend("topright",
       legend = paste("Mean =", round(mean(climate_similarity$climate_similarity), 1)),
       col = "red", lwd = 2, lty = 2)
```

## Step 6: Save Results

```{r save}
fwrite(climate_similarity, "../Data/Processed/climate_similarity.csv")
cat("Saved to Processed Data/climate_similarity.csv\n")
```

## Step 7: Merge into data_final.csv

The country pairs in `data_final.csv` are ordered by M49 numeric code, not alphabetically by ISO3. To match correctly, we create a canonical alphabetical key for both datasets before joining.

```{r merge}
data_final <- fread("../Data/Processed/data_final.csv")

# Remove climate_similarity column if it already exists (from a previous run)
if ("climate_similarity" %in% names(data_final)) {
  data_final[, climate_similarity := NULL]
}

climate_sim_merge <- climate_similarity %>%
  mutate(
    key_1 = pmin(iso3_1, iso3_2),
    key_2 = pmax(iso3_1, iso3_2)
  ) %>%
  select(key_1, key_2, climate_similarity)

data_final <- data_final %>%
  mutate(
    key_1 = pmin(iso3_o, iso3_d),
    key_2 = pmax(iso3_o, iso3_d)
  ) %>%
  left_join(climate_sim_merge, by = c("key_1", "key_2")) %>%
  select(-key_1, -key_2)

cat("Rows with climate data:", sum(!is.na(data_final$climate_similarity)),
    "out of", nrow(data_final), "\n")
cat("Missing:", sum(is.na(data_final$climate_similarity)), "\n")
```

Save the updated dataset:

```{r save-final}
fwrite(data_final, "../Data/Processed/data_final.csv")
cat("Updated data_final.csv saved\n")
```
