---
title: "Geographic Determinants of Food Similarity"
subtitle: "Geospatial Data Science -- BSE Spring 2026"
author: "Lisbeth Nordmeyer, Anabel Pichardo, Tianqin Tina Wang, Megan Yeo"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    latex_engine: pdflatex
header-includes:
  - \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

```{r packages, include=FALSE}
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(tidyr)
library(data.table)
library(tibble)
library(ggplot2)
library(viridis)
library(scales)
library(fixest)
library(countrycode)
library(knitr)
```

\newpage

# Introduction

Over the past decades, the global food landscape has been transformed by deepening economic integration. The spread of international trade networks, multinational food corporations, and global supply chains has made similar food products accessible across increasingly distant regions. Yet, persistent differences in consumption patterns suggest that geography continues to impose important constraints on what people eat. While a substantial body of research has examined how geography shapes diets within countries or specific regions, less is known about how it influences similarity in diets between countries on a global scale.

We address the following research question: *To what extent do geographic and socioeconomic factors explain dietary similarity between countries?* We conceptualize dietary similarity as a dyadic outcome between country pairs and investigate how it is influenced by geographic proximity (e.g., distance, shared borders), environmental similarity (e.g., climate zones, elevation, coastal access), as well as socioeconomic and historical factors such as GDP per capita differences, trade integration, and shared cultural or colonial ties.

To answer this question, we combine detailed food consumption data from the Food and Agriculture Organization (FAO) Food Balance Sheets with global datasets on geography, climate, trade, and economic development. We construct a bilateral food similarity index based on consumption shares across food categories and estimate gravity-style regressions to identify the key determinants of similarity. Our empirical analysis covers two points in time, 2010 and 2023, allowing us to assess both cross-sectional patterns and changes over time.

Our findings indicate that geography remains a strong and robust determinant of dietary similarity: countries that are geographically closer, share borders, or exhibit similar environmental conditions tend to have significantly more similar diets. At the same time, socioeconomic factors, particularly trade integration and shared historical ties, also play an important role, suggesting that globalization reinforces, but does not fully override, geographic patterns. Overall, the results highlight that while global diets may be converging, geography continues to exert a persistent influence on what people eat.

# Literature Review

Research has long documented the fact that geography has a significant impact on people's diets. For example, the Eurobarometer survey found that "proximity to the sea strongly influences higher consumption of fish and seafood" (European Commission, 2025), while a study on the Tibetan plateau highlights that the high altitude of the area results in the traditional diet being centered around livestock (Wang et al., 2023). There are also well-documented similarities in food by region, such as countries like France, Italy, Spain and Greece tending to adhere to Mediterranean diets (Mertens et al., 2018).

At the same time, there has been some literature documenting the relationship between geography and similarity in diet between countries. Zhu et al. (2013) analyze geography and the regional similarity of food in China, while Mertens et al. (2018) document similarity between selected European countries. We note that most of these studies have done so on a regional or country-specific level; our paper aims to do this analysis on a global scale.

In our paper, we investigate a few geographic variables that are likely significant determinants of diets: distance between countries, shared borders, (similarity in) climate zones, elevation, and distance to water bodies.

# Data Description

We use the UN Food and Agriculture Organization's Food Balance Sheets, which contain country-level data on the "pattern of a country's food supply during a specified reference period", including the amount consumed for each category of food (e.g. "Wheat and products", "Groundnuts", "Bovine meat"). To capture evolution over time, we use both 2010 and 2023 (earliest and latest available, respectively) data.

| Variable | Data Source |
|:---------|:------------|
| Food consumption shares | FAO Food Balance Sheets |
| Distance between countries | CEPII GeoDist + calculated ourselves |
| Population-weighted elevation | WorldPop population grid + R `elevatr` package |
| Climate zones | Köppen-Geiger Climate Zones raster (Beck et al., 2023) |
| Distance to coastline | Natural Earth coastline |
| Gravity controls (language, religion, colonial ties, FTA) | CEPII Gravity V2022 |
| Bilateral trade volume | CEPII BACI HS22 V2026 |
| GDP per capita (PPP) | World Bank |

# Methodology

## Food Similarity Index

We draw heavily from the methodology of the Country Food Similarity Index (ObjectiveLists, 2024) in calculating food similarity scores. We use the Finger-Kreinin Index commonly employed in trade literature to measure similarity in trade baskets (Finger and Kreinin, 1979). For each country, we first calculate the share of each category of food as a proportion of total consumption. For each country pair $j$ and $k$, we then calculate the overlapping share index as follows:

$$\text{Similarity}_{jk} = \sum_i \min(s_{i,j},\ s_{i,k})$$

where $s_{i,j}$ is consumption of food in category $i$ as a share of total consumption for country $j$, and $s_{i,k}$ is consumption of category $i$ as a share of total consumption for country $k$.

For instance, if "Wheat and products" constitutes a 0.01 share of total consumption in Spain and 0.02 share in France, the overlapping share is 0.01. Shares of consumption are calculated based on energy (kcal/capita/day), which the UN FAO also frequently employs.

The similarity score is then calculated as a weighted sum of the overlapping shares by country pair. Following the methodology of the Country Food Similarity Index, we assign a 30%, 30% and 40% weighting respectively to meat and dairy, starch, and fruit and vegetables, which delivers the most realistic similarity score between countries.

We note a few limitations in our calculation of similarity: while our study generalizes food consumption on a national level, we acknowledge that within a country there can be large disparities in cuisine. Since the UN FAO data only exists by country, we are unable to capture these nuances. Furthermore, we are only able to determine food similarity based on consumption of raw ingredients: we acknowledge that other factors like recipes and preparation methods are not reflected in our study.

Based on a visual inspection, this methodology generally yields sensible results. As expected, for instance, countries in the Mediterranean region have very high similarity scores, as do countries in geographical regions like Southeast Asia.

## Regression Specification

We estimate cross-sectional gravity-type regressions to examine the determinants of bilateral food similarity across countries. The unit of observation is a country pair. All models are estimated separately for 2010 and 2023.

The dependent variable is a continuous index of food similarity between two countries. The main explanatory variables include bilateral trade intensity, geographic distance, and standard gravity controls capturing cultural and historical proximity, such as common language, common religion, colonial ties, and trade agreements.

To account for unobserved country-specific heterogeneity, we include country fixed effects for both members of each pair. Standard errors are clustered at the country level in both dimensions (i.e., by each country in the dyad), allowing for arbitrary correlation of errors across dyads sharing a common country.

# Data Construction

## Data Cleaning

We filter the FAO Food Balance Sheets for food supply in kcal/capita/day (Element Code 664), merge in food category codes (meat, starch, vegetal products), exclude beverages, and compute consumption shares within each country. Countries are filtered to UN M49 recognized countries only.

```{r data-cleaning, eval=FALSE}
food_data <- read_csv("FoodBalanceSheets_E_All_Data_(Normalized).csv")
codes_updated <- read_csv("codes_updated.csv")
codes_updated <- unique(codes_updated)
m49_country_area_codes <- read_csv("m49_country_area_codes.csv")

# Filter for 2010 and 2023, supply in kcal/day
food_data <- food_data %>%
  filter(Year == 2023 | Year == 2010) %>%
  filter(`Element Code` == 664)

# Merge food category classifications
food_data <- food_data %>%
  left_join(codes_updated, by = "Item Code") %>%
  filter(!is.na(category))

# Calculate shares of total energy consumption
food_data <- food_data %>%
  group_by(Area) %>%
  mutate(share = Value / sum(Value)) %>%
  ungroup()

food_data <- food_data %>%
  mutate(area_code = `Area Code (M49)`,
         element_code = `Element Code`,
         item_code = `Item Code`) %>%
  select(area_code, item_code, Area, Item, share, category)

food_data$area_code <- sub("^'", "", food_data$area_code)
food_data <- food_data %>%
  mutate(area_code = as.numeric(area_code))
```

## Food Similarity Calculation

For each country pair, we compute the weighted overlap index across three food categories.

```{r food-similarity, eval=FALSE}
group_similarity <- function(df, c1, c2, groupname) {
  d1 <- df %>% filter(area_code == c1, category == groupname)
  d2 <- df %>% filter(area_code == c2, category == groupname)

  merged <- full_join(d1, d2, by = "Item", suffix = c("_1", "_2")) %>%
    mutate(
      share_1 = replace_na(share_1, 0),
      share_2 = replace_na(share_2, 0)
    )

  sum(pmin(merged$share_1, merged$share_2))
}

food_data <- food_data %>%
  filter(area_code %in% m49_country_area_codes$area_code) %>%
  filter(category != "beverages")

countries <- unique(food_data$area_code)
pairs <- expand.grid(c1 = countries, c2 = countries) %>%
  filter(c1 < c2)

similarity_results <- pairs %>%
  rowwise() %>%
  mutate(
    sim_meat   = group_similarity(food_data, c1, c2, "meat"),
    sim_starch = group_similarity(food_data, c1, c2, "starch"),
    sim_fv     = group_similarity(food_data, c1, c2, "vegetal_products"),
    similarity = 100 * (0.30 * sim_meat + 0.30 * sim_starch + 0.40 * sim_fv)
  )
```

## Non-Geographic Variables

We merge food similarity scores with bilateral trade data (BACI for 2023, CEPII Gravity for 2010), gravity controls from 2021 (common language, religion, colonial ties, FTA/WTO), and World Bank GDP per capita. The absolute GDP per capita difference is computed as a proxy for income dissimilarity.

```{r non-geo-merge, eval=FALSE}
# BACI trade data (2023)
baci <- trade2023 %>%
  left_join(trade_country, by = c("i" = "country_code")) %>%
  rename(iso3_o = country_iso3, country_name_o = country_name,
         iso2_o = country_iso2) %>%
  left_join(trade_country, by = c("j" = "country_code")) %>%
  rename(iso3_d = country_iso3, country_name_d = country_name,
         iso2_d = country_iso2)

baci_bilateral <- baci %>%
  group_by(t, iso3_o, iso3_d) %>%
  summarise(trade_value = sum(v, na.rm = TRUE), .groups = "drop") %>%
  rename("year" = "t", "trade_total" = "trade_value")

# Gravity controls from 2021 (time-invariant)
gravity_controls <- gravity %>%
  filter(year == 2021, country_exists_o == 1, country_exists_d == 1) %>%
  select(iso3_o, iso3_d, comlang_off, comrelig, col45, comcol, fta_wto,
         scaled_sci_2021, dist, distcap) %>%
  rename_with(~ paste0(., "_2021"), -c(iso3_o, iso3_d, scaled_sci_2021))

# Harmonise FAO M49 codes to ISO3
similarity_merged <- bind_rows(
  similarity_2010 %>% mutate(year = 2010),
  similarity_2023 %>% mutate(year = 2023)
) %>%
  mutate(
    iso3_o = countrycode(c1, origin = "un", destination = "iso3c"),
    iso3_d = countrycode(c2, origin = "un", destination = "iso3c")
  ) %>%
  rename("m49_1" = "c1", "m49_2" = "c2")

# GDP long format
gdp_long <- gdp %>%
  select("Country Code", "2010", "2023") %>%
  pivot_longer(cols = c("2010", "2023"), names_to = "year",
               values_to = "gdp_pc") %>%
  rename(iso3 = "Country Code") %>%
  mutate(year = as.integer(year))

# Final merge: similarity + trade + gravity + GDP
data_final <- bind_rows(
  similarity_merged %>% filter(year == 2010) %>%
    left_join(trade_2010, by = c("iso3_o", "iso3_d")) %>%
    left_join(gravity_controls, by = c("iso3_o", "iso3_d")),
  similarity_merged %>% filter(year == 2023) %>%
    left_join(baci_bilateral %>% filter(year == 2023),
              by = c("iso3_o", "iso3_d", "year")) %>%
    left_join(gravity_controls, by = c("iso3_o", "iso3_d"))
) %>%
  left_join(gdp_long, by = c("iso3_o" = "iso3", "year")) %>%
  rename(gdp_pc_o = gdp_pc) %>%
  left_join(gdp_long, by = c("iso3_d" = "iso3", "year")) %>%
  rename(gdp_pc_d = gdp_pc) %>%
  mutate(gdp_pc_diff = abs(gdp_pc_o - gdp_pc_d))
```

## Geographic Variables

### Distance

We compute pairwise distances between countries using two methods: (1) point-on-surface of each country polygon, and (2) population-weighted centroids based on city populations from Natural Earth. The latter provides a better approximation of where economic activity is concentrated.

```{r geo-distance, eval=FALSE}
world_main <- ne_countries(scale = "medium", returnclass = "sf")
cities_main <- ne_download(scale = "medium", type = "populated_places",
                           category = "cultural", returnclass = "sf") %>%
  rename_with(tolower)

world <- world_main %>%
  filter(!is.na(adm0_a3), adm0_a3 != "-99") %>%
  select(adm0_a3, geometry) %>%
  group_by(adm0_a3) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

# Population-weighted centroids
coords <- st_coordinates(cities_main)

pop_centroids <- cities_main %>%
  mutate(x = coords[, 1], y = coords[, 2]) %>%
  st_drop_geometry() %>%
  filter(!is.na(adm0_a3), adm0_a3 != "-99", !is.na(pop_max)) %>%
  group_by(adm0_a3) %>%
  summarise(
    x = weighted.mean(x, pop_max, na.rm = TRUE),
    y = weighted.mean(y, pop_max, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_as_sf(coords = c("x", "y"), crs = 4326)

# Pairwise distance matrix
pop_dist_matrix <- st_distance(pop_centroids)
pop_dist_matrix <- drop_units(set_units(pop_dist_matrix, "km"))
rownames(pop_dist_matrix) <- pop_centroids$adm0_a3
colnames(pop_dist_matrix) <- pop_centroids$adm0_a3

pop_dist_df <- as.data.frame(pop_dist_matrix) %>%
  rownames_to_column("iso3_o") %>%
  pivot_longer(cols = -iso3_o, names_to = "iso3_d",
               values_to = "pop_dist_km") %>%
  filter(iso3_o != iso3_d)
```

### Shared Borders

We use `st_touches()` to identify contiguous country pairs and create a binary adjacency indicator.

```{r geo-borders, eval=FALSE}
border_mat <- st_touches(world, sparse = FALSE)
dimnames(border_mat) <- list(world$adm0_a3, world$adm0_a3)

border_df <- as.data.frame(border_mat) %>%
  rownames_to_column("iso3_o") %>%
  pivot_longer(cols = -iso3_o, names_to = "iso3_d",
               values_to = "shared_border") %>%
  mutate(shared_border = as.integer(shared_border)) %>%
  filter(iso3_o != iso3_d)
```

### Coastal Access

We compute each country's coastal status (1/0) and shortest distance to the nearest coastline. Coastal countries are assigned distance = 0.

```{r geo-coast, eval=FALSE}
coastline_main <- ne_download(scale = "medium", type = "coastline",
                              category = "physical", returnclass = "sf")

world_coast <- st_transform(world, 4326)
coastline_coast <- st_make_valid(coastline_main) %>% st_transform(4326)
coast_union <- st_union(coastline_coast)

coastal_dummy <- lengths(st_intersects(world_coast, coast_union)) > 0
dist_to_coast <- st_distance(world_coast, coast_union)

coast_df <- world_coast %>%
  st_drop_geometry() %>%
  mutate(
    iso3 = adm0_a3,
    coastal = as.integer(coastal_dummy),
    dist_coast_km = as.numeric(set_units(dist_to_coast, "km"))
  ) %>%
  select(iso3, coastal, dist_coast_km) %>%
  mutate(dist_coast_km = if_else(coastal == 1, 0, dist_coast_km))
```

### Climate Similarity

We use the Köppen-Geiger climate classification raster from Beck et al. (2023), which classifies every pixel on Earth into one of 30 climate zones. For each country, we compute the fraction of its land area falling in each zone, using `exact_extract()` with coverage fractions to handle partial border pixels. Climate similarity between country pairs is then computed using the same Finger-Kreinin overlap method as the food similarity index:

$$\text{Climate Similarity}_{jk} = \sum_{z=1}^{30} \min(s_{jz},\ s_{kz}) \times 100$$

where $s_{jz}$ is country $j$'s share of land area in climate zone $z$.

```{r geo-climate, eval=FALSE}
library(terra)
library(exactextractr)

# Köppen-Geiger raster (1991-2020, 0.1 degree resolution)
koppen_raster <- rast("../Data/Raw/1991_2020/koppen_geiger_0p1.tif")
world <- ne_countries(scale = "medium", returnclass = "sf")

# Extract raster pixels per country
climate_extract <- exact_extract(koppen_raster, world, include_cols = "iso_a3")
climate_df <- bind_rows(climate_extract)
names(climate_df)[names(climate_df) == "value"] <- "climate_zone"

# Share of each climate zone per country
climate_shares <- climate_df %>%
  filter(!is.na(climate_zone), climate_zone != 0) %>%
  group_by(iso_a3, climate_zone) %>%
  summarise(area = sum(coverage_fraction), .groups = "drop") %>%
  group_by(iso_a3) %>%
  mutate(share = area / sum(area)) %>%
  ungroup() %>%
  select(iso_a3, climate_zone, share) %>%
  filter(!is.na(iso_a3), iso_a3 != "-99")

# Pivot to wide matrix for vectorised computation
climate_wide <- climate_shares %>%
  pivot_wider(id_cols = iso_a3, names_from = climate_zone,
              values_from = share, values_fill = 0)

country_ids <- climate_wide$iso_a3
climate_mat <- as.matrix(climate_wide[, -1])
rownames(climate_mat) <- country_ids

n <- length(country_ids)
pairs <- combn(n, 2)

# Overlap similarity for each pair
similarity_vec <- sapply(seq_len(ncol(pairs)), function(k) {
  sum(pmin(climate_mat[pairs[1, k], ], climate_mat[pairs[2, k], ]))
})

climate_similarity <- data.frame(
  iso3_1 = country_ids[pairs[1, ]],
  iso3_2 = country_ids[pairs[2, ]],
  climate_similarity = similarity_vec * 100
)
```

### Elevation Similarity

We calculate the average elevation in each country, weighted by population, using a global population raster from WorldPop and an elevation raster from the R `elevatr` package. We then compute the bilateral similarity as the negative log of the absolute difference in population-weighted elevation between country pairs.

```{r geo-elevation, eval=FALSE}
library(raster)
library(elevatr)

population <- raster("global_pop.tif")
countries <- read_sf("ne_10m_admin_0_countries/ne_10m_admin_0_countries.shp")

# Calculate elevation raster
elevation <- get_elev_raster(countries, z = 5)
countries_shape <- vect(countries)
elevation_mask <- mask(elevation, countries)

# Reproject and compute weighted mean elevation
population <- rast(population)
elevation_mask <- rast(elevation_mask)
countries <- st_transform(countries, crs(elevation_mask))
population <- project(population, elevation_mask)

weighted <- population * elevation_mask
weighted_sum <- zonal(weighted, countries_shape, fun = "sum", na.rm = TRUE)
pop_sum <- zonal(population, countries_shape, fun = "sum", na.rm = TRUE)
weighted_mean <- weighted_sum / pop_sum
countries$weighted_elevation <- weighted_mean$global_pop

# Bilateral elevation difference
data_final <- data_final %>%
  left_join(countries_o, by = "iso3_o") %>%
  left_join(countries_d, by = "iso3_d") %>%
  mutate(elevation_similarity = -log(abs(weighted_elevation_o -
                                          weighted_elevation_d)))
```

### Final Dataset Assembly

All geographic variables are merged into a single bilateral panel dataset.

```{r geo-final-merge, eval=FALSE}
data_final_geo <- data_final %>%
  left_join(dist_df, by = c("iso3_o", "iso3_d")) %>%
  mutate(log_dist_centroid = log(dist_km)) %>%
  left_join(pop_dist_df, by = c("iso3_o", "iso3_d")) %>%
  mutate(log_dist_pop_centroid = log(pop_dist_km)) %>%
  left_join(border_df, by = c("iso3_o", "iso3_d")) %>%
  mutate(shared_border = replace_na(shared_border, 0)) %>%
  left_join(coast_df, by = c("iso3_o" = "iso3")) %>%
  rename(coastal_o = coastal, dist_coast_km_o = dist_coast_km) %>%
  left_join(coast_df, by = c("iso3_d" = "iso3")) %>%
  rename(coastal_d = coastal, dist_coast_km_d = dist_coast_km) %>%
  mutate(diff_dist_coast_km = abs(dist_coast_km_o - dist_coast_km_d))
```

\newpage

# Results

```{r load-precomputed, echo=FALSE}
# Load pre-computed datasets for visualisations and regressions
data_final_geo <- fread("../Data/Processed/data_final_geo.csv")

climate_similarity <- fread("../Data/Processed/climate_similarity.csv")

food_data_2023 <- fread("../Data/Processed/food_data.csv")
food_data_2010 <- fread("../Data/Processed/food_data_2010.csv")
food_merged <- bind_rows(
  food_data_2010 %>% mutate(year = 2010),
  food_data_2023 %>% mutate(year = 2023)
) %>%
  mutate(iso3c = countrycode(area_code, origin = "un", destination = "iso3c"))

similarity_2010 <- fread("../Data/Processed/similarity_results_2010.csv")
similarity_2023 <- fread("../Data/Processed/similarity_results.csv")
similarity_merged <- bind_rows(
  similarity_2010 %>% mutate(year = 2010),
  similarity_2023 %>% mutate(year = 2023)
) %>%
  mutate(
    iso3_o = countrycode(c1, origin = "un", destination = "iso3c"),
    iso3_d = countrycode(c2, origin = "un", destination = "iso3c")
  ) %>%
  rename("m49_1" = "c1", "m49_2" = "c2")

countries_elev <- read_sf("../Data/Processed/countries_elevation.shp")

world_main <- ne_countries(scale = "medium", returnclass = "sf")
world_map <- world_main %>% select(adm0_a3, geometry)
```

## Descriptive Visualisations

### Average Food Similarity by Country

The following choropleth maps show each country's mean pairwise food similarity with all other countries, faceted by year.

```{r map-avg-food, fig.width=10, fig.height=5}
similarity_map <- data_final_geo %>%
  group_by(iso3_o, year) %>%
  summarise(mean_similarity = mean(similarity, na.rm = TRUE), .groups = "drop")

map_data <- world_map %>%
  left_join(similarity_map, by = c("adm0_a3" = "iso3_o"))

plot_data <- map_data %>% filter(year %in% c(2010, 2023))

ggplot(data = plot_data) +
  geom_sf(data = world_map, fill = "grey80", color = "white",
          linewidth = 0.1, inherit.aes = FALSE) +
  geom_sf(aes(fill = mean_similarity), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80",
                       name = "Avg. Food\nSimilarity") +
  facet_wrap(~ year) +
  labs(title = "Average Food Similarity by Country: 2010 vs 2023",
       caption = "Source: Own calculations based on FAO Food Balance Sheets") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "bottom")
```

Food similarity is strongly regional in both years: Europe, North America, and parts of East Asia show relatively high similarity, while Sub-Saharan Africa and parts of Southeast Asia remain more distinct.

From 2010 to 2023, there is a mild increase in similarity in several regions, suggesting some dietary convergence over time. However, the global pattern remains uneven, as low-similarity clusters persist, especially in Africa. This indicates that geography and regional factors still play a major role in shaping diets. An interactive version of this map is available at https://apichardo1997.github.io/food-atlas/ (Food Similarity tab).

### Food Similarity to China

The following maps show how similar each country's diet is to China's, capped at the maximum non-China value so color variation among other countries is visible.

```{r map-china, fig.width=10, fig.height=5}
china_similarity <- similarity_merged %>%
  filter(year %in% c(2010, 2023), iso3_o == "CHN" | iso3_d == "CHN") %>%
  mutate(iso3 = if_else(iso3_o == "CHN", iso3_d, iso3_o)) %>%
  select(iso3, year, similarity)

china_similarity <- bind_rows(
  china_similarity,
  tibble(iso3 = "CHN", year = c(2010, 2023), similarity = 100)
)

map_china <- world_map %>%
  left_join(china_similarity, by = c("adm0_a3" = "iso3")) %>%
  filter(year %in% c(2010, 2023))

world_map_bg <- bind_rows(
  world_map %>% mutate(year = 2010),
  world_map %>% mutate(year = 2023)
)

non_china_max <- map_china %>%
  filter(adm0_a3 != "CHN", !is.na(similarity)) %>%
  summarise(max_sim = max(similarity)) %>%
  pull(max_sim)

ggplot() +
  geom_sf(data = world_map_bg, fill = "grey80", color = "white",
          linewidth = 0.1, inherit.aes = FALSE) +
  geom_sf(data = map_china, aes(fill = similarity), color = "white",
          linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma", limits = c(0, non_china_max),
                       oob = squish, na.value = "grey80",
                       name = "Similarity\nto China") +
  facet_wrap(~ year) +
  labs(title = "Food Similarity to China: 2010 vs 2023",
       caption = "Source: Own calculations based on FAO Food Balance Sheets") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "bottom")
```

The maps reveal high similarity to China in East and Southeast Asia, as expected, but also notable similarity in parts of Europe, North America, and some South American countries such as Peru and Guyana. While the Asian pattern reflects geographic proximity and shared dietary staples, the similarities observed in regions like Europe and North America likely capture more globalized consumption patterns. In the case of Peru, the relatively high similarity is consistent with historical Chinese migration and its lasting influence on local diets, although the measure primarily reflects overlap in broad food categories rather than specific cuisines. An interactive version is available at https://apichardo1997.github.io/food-atlas/ (Food Similarity tab).

### Seafood Share of Diet

The following map highlights how dependent countries are on fish and other seafood as a share of total caloric intake.

```{r map-seafood, fig.width=9, fig.height=5}
seafood_codes <- c(2761, 2762, 2763, 2764, 2765, 2766, 2767)

seafood_shares <- food_merged %>%
  filter(year == 2023, item_code %in% seafood_codes) %>%
  group_by(iso3c) %>%
  summarise(seafood_share = sum(share, na.rm = TRUE), .groups = "drop")

map_seafood <- world_main %>%
  select(adm0_a3, geometry) %>%
  left_join(seafood_shares, by = c("adm0_a3" = "iso3c"))

ggplot(map_seafood) +
  geom_sf(aes(fill = seafood_share * 100), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "mako", na.value = "grey80",
                       name = "Seafood Share\n(% of diet)", transform = "sqrt") +
  labs(title = "Seafood Share of Total Diet (2023)",
       subtitle = "Share of total caloric intake from seafood products",
       caption = "Source: FAO Food Balance Sheets") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "right")
```

As expected, seafood consumption is highest in coastal and island regions, particularly in Southeast Asia and parts of Northern Europe, while inland regions, especially in Africa and Central Asia, show very low shares. An interactive version is available at https://apichardo1997.github.io/food-atlas/ (Geography tab).

### Climate Similarity

Average climate similarity per country, computed as the mean Finger-Kreinin overlap in Köppen-Geiger climate zone distributions across all partners.

```{r map-avg-climate, fig.width=10, fig.height=6}
world_plot <- world_main %>%
  filter(iso_a3 != "ATA") %>%
  st_transform(crs = "+proj=robin")

avg_sim <- bind_rows(
  climate_similarity %>% select(iso3 = iso3_1, climate_similarity),
  climate_similarity %>% select(iso3 = iso3_2, climate_similarity)
) %>%
  group_by(iso3) %>%
  summarise(avg_climate_sim = mean(climate_similarity, na.rm = TRUE))

world_avg <- world_plot %>%
  left_join(avg_sim, by = c("iso_a3" = "iso3"))

ggplot(world_avg) +
  geom_sf(aes(fill = avg_climate_sim), color = "grey30", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "magma", na.value = "grey90",
    name = "Avg. Climate\nSimilarity",
    breaks = seq(10, 50, 10)
  ) +
  labs(
    title = "Average Climate Similarity by Country",
    subtitle = "Mean Köppen-Geiger overlap with all other countries (0-100 scale)",
    caption = "Source: Beck et al. (2023). Own calculations."
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9, color = "grey40", hjust = 0.5,
                                 margin = margin(b = 10)),
    plot.caption = element_text(size = 7, color = "grey50", hjust = 1),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    legend.key.height = unit(0.3, "cm")
  )
```

Countries near the tropics tend to have higher average climate similarity, as tropical climate zones are widespread and shared across many equatorial nations. In contrast, countries with more unique climate profiles, such as Nordic countries or desert states, score lower. This pattern confirms that the Köppen-Geiger overlap index captures meaningful variation in environmental conditions across the globe. Interactive versions of these maps, with country-level hover details, are available at https://apichardo1997.github.io/food-atlas/ (Climate tab).

The following map shows climate similarity to Spain, illustrating how the overlap index captures intuitive Mediterranean patterns.

```{r map-spain-climate, fig.width=10, fig.height=6}
spain_sim <- climate_similarity %>%
  filter(iso3_1 == "ESP" | iso3_2 == "ESP") %>%
  mutate(partner = ifelse(iso3_1 == "ESP", iso3_2, iso3_1)) %>%
  select(partner, climate_similarity)

spain_sim <- bind_rows(spain_sim,
                       data.frame(partner = "ESP", climate_similarity = 100))

world_spain <- world_plot %>%
  left_join(spain_sim, by = c("iso_a3" = "partner"))

ggplot(world_spain) +
  geom_sf(aes(fill = climate_similarity), color = "grey30", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "inferno", na.value = "grey90",
    name = "Climate\nSimilarity\nto Spain",
    limits = c(0, 100)
  ) +
  labs(
    title = "Climate Similarity to Spain",
    subtitle = "Overlap in Köppen-Geiger climate zone distributions (0-100)",
    caption = "Source: Beck et al. (2023). Own calculations."
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9, color = "grey40", hjust = 0.5,
                                 margin = margin(b = 10)),
    plot.caption = element_text(size = 7, color = "grey50", hjust = 1),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    legend.key.height = unit(0.3, "cm")
  )
```

Spain's highest climate similarity is with other Mediterranean and semi-arid countries, including Morocco, Tunisia, Italy, and parts of the Middle East. The pattern confirms that the index captures intuitive geographic clusters. Countries with no climate overlap with Spain, such as equatorial and polar nations, appear in the lowest category.

### Population-Weighted Elevation

The following map shows the average elevation experienced by each country's population, weighted by population density. An interactive version is available at https://apichardo1997.github.io/food-atlas/ (Elevation tab).

```{r map-elevation, fig.width=10, fig.height=5}
ggplot(data = countries_elev) +
  geom_sf(fill = "grey80", color = "white", linewidth = 0.1,
          inherit.aes = FALSE) +
  geom_sf(aes(fill = log(wghtd_l)), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80",
                       name = "Log Pop-Weighted\nAvg. Elevation") +
  labs(title = "Log Population-Weighted Average Elevation",
       caption = "Source: Own calculations, R elevatr package + WorldPop") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "bottom")
```

Countries with the highest population-weighted elevation include Bolivia, Ethiopia, and parts of Central Asia, where populations are concentrated at high altitudes. Low-elevation countries are concentrated in coastal Europe, Southeast Asia, and island nations.

### Average Elevation Similarity

```{r map-elev-similarity, fig.width=10, fig.height=5}
elevation_similarity_map <- data_final_geo %>%
  filter(year == 2010) %>%
  group_by(iso3_o) %>%
  summarise(mean_similarity = mean(elevation_similarity, na.rm = TRUE),
            .groups = "drop")

map_data_elev <- world_map %>%
  left_join(elevation_similarity_map, by = c("adm0_a3" = "iso3_o"))

ggplot(data = map_data_elev) +
  geom_sf(data = world_map, fill = "grey80", color = "white",
          linewidth = 0.1, inherit.aes = FALSE) +
  geom_sf(aes(fill = mean_similarity), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80",
                       name = "Avg. Elevation\nSimilarity") +
  labs(title = "Average Population-Weighted Elevation Similarity by Country",
       caption = "Source: Own calculations, R elevatr package + WorldPop") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "bottom")
```

Lower values indicate a greater average elevation difference from other countries. Countries at extreme elevations, such as Bolivia and Nepal, show the lowest average similarity, while countries at moderate elevations near the global mean tend to score higher.

### Distribution of Food Similarity

```{r dist-overall, fig.width=8, fig.height=4}
ggplot(data_final_geo, aes(x = similarity, fill = factor(year),
                            color = factor(year))) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("2010" = "#2166ac", "2023" = "#d6604d"),
                    name = "Year") +
  scale_color_manual(values = c("2010" = "#2166ac", "2023" = "#d6604d"),
                     name = "Year") +
  labs(title = "Distribution of Pairwise Food Similarity",
       x = "Food Similarity", y = "Density",
       caption = "Source: Own calculations based on FAO Food Balance Sheets") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

The distribution of pairwise food similarity shifts slightly to the right from 2010 to 2023, indicating a modest overall increase in similarity between countries. The 2023 distribution shows fewer very low-similarity pairs and a higher concentration around mid-to-high similarity values, suggesting gradual convergence in global diets.

### Distribution by Food Category

```{r dist-categories, fig.width=10, fig.height=4}
data_final_geo %>%
  select(year, sim_meat, sim_starch, sim_fv) %>%
  pivot_longer(cols = -year, names_to = "category", values_to = "similarity") %>%
  mutate(category = recode(category,
                           sim_meat   = "Meat",
                           sim_starch = "Starch",
                           sim_fv     = "Fruits & Vegetables")) %>%
  ggplot(aes(x = similarity, fill = factor(year), color = factor(year))) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("2010" = "#2166ac", "2023" = "#d6604d"),
                    name = "Year") +
  scale_color_manual(values = c("2010" = "#2166ac", "2023" = "#d6604d"),
                     name = "Year") +
  facet_wrap(~ category) +
  labs(title = "Distribution of Food Similarity by Category",
       x = "Similarity", y = "Density",
       caption = "Source: Own calculations based on FAO Food Balance Sheets") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

The distribution for meat shows the clearest rightward shift from 2010 to 2023, indicating higher similarity across countries in this category, whereas fruits & vegetables remain almost unchanged and starch shows only a modest shift. This suggests that the overall increase in food similarity is driven primarily by convergence in meat consumption patterns rather than by the other categories.

\newpage

## Regression Analysis

We estimate six specifications: cross-sections for 2010 and 2023, with and without country fixed effects, and with an additional GDP per capita difference term. All models include climate similarity and elevation similarity alongside the baseline gravity variables.

```{r reg-data-prep}
reg_data <- data_final_geo %>%
  mutate(
    year = as.integer(year),

    # Trade
    trade_total = replace_na(trade_total, 0),
    log_trade = log1p(trade_total),

    # Distance measures
    log_pop_dist = if_else(
      !is.na(pop_dist_km) & pop_dist_km > 0,
      log(pop_dist_km),
      NA_real_
    ),
    log_diff_dist_coast_km = log1p(diff_dist_coast_km),

    # GDP difference
    gdp_pc_diff = if_else(
      is.na(gdp_pc_diff) & !is.na(gdp_pc_o) & !is.na(gdp_pc_d),
      abs(gdp_pc_o - gdp_pc_d),
      gdp_pc_diff
    ),
    log_gdp_diff = if_else(
      !is.na(gdp_pc_diff) & gdp_pc_diff > 0,
      log(gdp_pc_diff),
      NA_real_
    ),

    # Make sure controls are numeric
    shared_border = as.numeric(shared_border),
    fta_wto_2021 = as.numeric(fta_wto_2021),
    comlang_off_2021 = as.numeric(comlang_off_2021),
    comrelig_2021 = as.numeric(comrelig_2021),
    comcol_2021 = as.numeric(comcol_2021),
    col45_2021 = as.numeric(col45_2021),

    # Additional geographic variables
    climate_similarity = as.numeric(climate_similarity),
    elevation_similarity = as.numeric(elevation_similarity)
  )
```

```{r regressions, results='asis'}
# (1) 2010: no country fixed effects
m2010_clim_elev <- feols(
  similarity ~ log_pop_dist + log_diff_dist_coast_km + shared_border +
    climate_similarity + elevation_similarity + log_trade +
    fta_wto_2021 + comlang_off_2021 + comrelig_2021 +
    comcol_2021 + col45_2021,
  data = reg_data %>% filter(year == 2010),
  vcov = ~iso3_o + iso3_d
)

# (2) 2023: no country fixed effects
m2023_clim_elev <- feols(
  similarity ~ log_pop_dist + log_diff_dist_coast_km + shared_border +
    climate_similarity + elevation_similarity + log_trade +
    fta_wto_2021 + comlang_off_2021 + comrelig_2021 +
    comcol_2021 + col45_2021,
  data = reg_data %>% filter(year == 2023),
  vcov = ~iso3_o + iso3_d
)

# (3) 2010: with country fixed effects
m2010_fe_clim_elev <- feols(
  similarity ~ log_pop_dist + log_diff_dist_coast_km + shared_border +
    climate_similarity + elevation_similarity + log_trade +
    fta_wto_2021 + comlang_off_2021 + comrelig_2021 +
    comcol_2021 + col45_2021 | iso3_o + iso3_d,
  data = reg_data %>% filter(year == 2010),
  vcov = ~iso3_o + iso3_d
)

# (4) 2023: with country fixed effects
m2023_fe_clim_elev <- feols(
  similarity ~ log_pop_dist + log_diff_dist_coast_km + shared_border +
    climate_similarity + elevation_similarity + log_trade +
    fta_wto_2021 + comlang_off_2021 + comrelig_2021 +
    comcol_2021 + col45_2021 | iso3_o + iso3_d,
  data = reg_data %>% filter(year == 2023),
  vcov = ~iso3_o + iso3_d
)

# (5) 2010: with country fixed effects and GDP difference
m2010_fe_gdp_clim_elev <- feols(
  similarity ~ log_pop_dist + log_diff_dist_coast_km + shared_border +
    climate_similarity + elevation_similarity + log_gdp_diff +
    log_trade + fta_wto_2021 + comlang_off_2021 +
    comrelig_2021 + comcol_2021 + col45_2021 | iso3_o + iso3_d,
  data = reg_data %>% filter(year == 2010),
  vcov = ~iso3_o + iso3_d
)

# (6) 2023: with country fixed effects and GDP difference
m2023_fe_gdp_clim_elev <- feols(
  similarity ~ log_pop_dist + log_diff_dist_coast_km + shared_border +
    climate_similarity + elevation_similarity + log_gdp_diff +
    log_trade + fta_wto_2021 + comlang_off_2021 +
    comrelig_2021 + comcol_2021 + col45_2021 | iso3_o + iso3_d,
  data = reg_data %>% filter(year == 2023),
  vcov = ~iso3_o + iso3_d
)

etable(
  m2010_clim_elev,
  m2023_clim_elev,
  m2010_fe_clim_elev,
  m2023_fe_clim_elev,
  m2010_fe_gdp_clim_elev,
  m2023_fe_gdp_clim_elev,
  fitstat = ~n + r2,
  tex = TRUE
) |> knitr::asis_output()
```

# Discussion

Overall, the findings provide strong and consistent evidence that geographic and environmental factors play a central role in shaping dietary similarity between countries. Geographic distance emerges as one of the most robust determinants. Across all specifications, the coefficient on the logarithm of population-weighted distance is negative and highly significant, indicating that countries that are geographically closer tend to have more similar diets. This result is consistent with standard gravity-type findings and supports the idea that proximity facilitates similarities in food availability, agricultural conditions, and consumption patterns.

We also find evidence that geographic features beyond simple distance matter. Differences in access to the coast are negatively associated with dietary similarity, suggesting that countries with similar proximity to the sea tend to exhibit more similar consumption patterns, likely reflecting the importance of seafood and maritime trade. While this effect weakens somewhat once country fixed effects are included, it remains economically meaningful. Similarly, sharing a common border is associated with greater dietary similarity, although this effect becomes statistically significant only after controlling for country fixed effects, indicating that neighboring countries share similarities beyond general geographic proximity.

Environmental similarity plays a particularly strong and consistent role. Both climate similarity and elevation similarity are positively and highly significant across all specifications. Countries that share similar climate zones or elevation profiles tend to have significantly more similar diets, even after controlling for country fixed effects and other covariates. These findings highlight the importance of environmental conditions in shaping agricultural production and food availability, and constitute one of the key contributions of this paper.

Turning to economic factors, bilateral trade is positively associated with dietary similarity in models without fixed effects, but loses statistical significance once country fixed effects are included. This suggests that the observed relationship between trade and dietary similarity may be driven by underlying country characteristics or reverse causality, rather than a direct causal effect. Differences in GDP per capita appear to play a role, but do not substantially alter the main geographic coefficients, indicating that income differences are of secondary importance compared to geographic and environmental factors.

Cultural and historical variables show more mixed results. Shared language and religion are positively associated with dietary similarity in simpler specifications, but their effects become weaker or insignificant once fixed effects are included. In contrast, shared colonial ties and trade agreements remain positively associated with dietary similarity, with trade agreements showing a particularly strong and robust effect across all models.

Importantly, the estimated coefficients are remarkably stable between 2010 and 2023, suggesting that the role of geography and environmental conditions in shaping dietary similarity has remained persistent over time, despite increasing globalization.

# Conclusion

Taken together, these results suggest that while globalization and economic integration contribute to dietary convergence, they do not fully offset the influence of geography. Instead, geographic proximity and environmental similarity remain the dominant forces shaping dietary patterns across countries. This finding complements existing evidence on the global "nutrition transition" (Traill et al., 2014), suggesting that while diets may be converging in some dimensions, fundamental geographic and environmental constraints continue to play a persistent role.

# References

- Beck, H. E., et al. (2023). High-resolution (1 km) Köppen-Geiger maps for 1901--2099 based on constrained CMIP6 projections. *Scientific Data*.
- European Commission (2025). Eurobarometer survey on fishery and aquaculture products consumption.
- Finger, J. M., & Kreinin, M. E. (1979). A Measure of 'Export Similarity' and Its Possible Uses. *The Economic Journal*, 89(356), 905--912.
- ObjectiveLists (2024). Country Food Similarity Index. https://objectivelists.com/country-food-similarity-index/
- Mertens, E., et al. (2018). Dietary patterns across European countries. *Nutrients*. https://pmc.ncbi.nlm.nih.gov/articles/PMC6561990/
- Traill, W. B., et al. (2014). Nutrition transition and globalization. *Nutrition Reviews*. https://doi.org/10.1111/nure.12134
- Wang, Y., et al. (2023). Dietary patterns on the Tibetan Plateau. https://pmc.ncbi.nlm.nih.gov/articles/PMC10181060/
- Zhu, C., et al. (2013). Regional food similarity in China. *PLoS ONE*. https://pmc.ncbi.nlm.nih.gov/articles/PMC3832477/
